---
title: "Assignment: Among Site (Beta) Diversity"
author: "Michelle Benavidez; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this Assignment, we move beyond the investigation of within-site $\alpha$-diversity.
We will explore $\beta$-diversity, which is defined as the diversity that occurs among sites.
This requires that we examine the compositional similarity of assemblages that vary in space or time.

After completing this exercise you will know how to:

1. formally quantify $\beta$-diversity
2. visualize $\beta$-diversity with heatmaps, cluster analysis, and ordination
3. test hypotheses about $\beta$-diversity using multivariate statistics

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the Handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the Handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. When you are done wit the Assignment, **Knit** the text and code into a html file.
7. After Knitting, please submit the completed Assignment by creating a **pull request** via GitHub.
Your pull request should include this file *beta_assignment.Rmd* and the html output of `Knitr` (*beta_assignment.html*).

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/Week3-Beta*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}
rm(list=ls())
getwd()
setwd("~/GitHub/QB2017_Benavidez/Week3-Beta")
require("vegan")
```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `doubs` dataset from the `ade4` package, and
2. explore the structure of the dataset.

```{r}
data(doubs)
str(doubs, max=1)
head(doubs$env)
```

***Question 1***:  Describe some of the attributes of the `doubs` dataset.  

a.  How many objects are in `doubs`?
b.  How many fish species are there in the `doubs` dataset? 
c.  How many sites are in the `doubs` dataset? 

> Answer 1a***: Four objects
> Answer 1b***: 27 fish species 
> Answer 1c***: 30 sites

### Visualizing the Doubs River Dataset

***Question 2***: Answer the following questions based on the spatial patterns of richness (i.e., $\alpha$-diversity) and Brown Trout (*Salmo trutta*) abundance in the Doubs River.  

a.  How does fish richness vary along the sampled reach of the Doubs River?
b.  How does Brown Trout (*Salmo trutta*) abundance vary along the sampled reach of the Doubs River?
c.  What do these patterns say about the limitations of using richness when examining patterns of biodiversity?

> Answer 2a***:  Richness seems to decease as you move from downstream to upstream.
> Answer 2b***:  Abundance of brown trout seems to increase as you move from downstream to upstream.
> Answer 2c***:  Examining richness alones many be limited because it does a poor job of describing patterns in abundace.  Downstream you have many species with low abundances while upstream you have few species at high abundances.

## 3) QUANTIFYING BETA-DIVERSITY

In the R code chunk below, do the following:  

1. write a function (`beta.w()`) to calculate Whittaker's $\beta$-diversity (i.e., $\beta_{w}$) that accepts a site-by-species matrix with optional arguments to specify pairwise turnover between two sites, and
2. use this function to analyze various aspects of $\beta$-diversity in the Doubs River.  

```{r}
beta.w=function(site.by.species = ""){
  SbyS.pa=decostand(site.by.species, method = "pa") 
  S=ncol(SbyS.pa[,which(colSums(SbyS.pa) > 0)]) 
  a.bar <- mean(specnumber(SbyS.pa))
  b.w=round(S/a.bar, 3)
  return(b.w)
}

beta.w.turnover=function(site.by.species="",sitenum1="",sitenum2= "",pairwise= FALSE){
  if (pairwise == TRUE){
    if (sitenum1 == "" | sitenum2 == "") {
      print("Error: please specify sites to compare") 
      return(NA)}
    site1 = site.by.species[sitenum1,]
    site2 = site.by.species[sitenum2,]
    site1 = subset(site1, select = site1 > 0)
    site2 = subset(site2, select = site2 > 0)
    gamma = union(colnames(site1), colnames(site2))
    s = length(gamma)
    a.bar = mean(c(specnumber(site1), specnumber(site2))) 
    b.w = round(s/a.bar - 1, 3)
    return(b.w)
  }
else{
  SbyS.pa <- decostand(site.by.species, method = "pa")
  S <- ncol(SbyS.pa[,which(colSums(SbyS.pa) > 0)]) 
  a.bar <- mean(specnumber(SbyS.pa)) 
  b.w <- round(S/a.bar, 3)
  return(b.w)
  }
}
```

***Question 3***: Using your `beta.w()` function above, answer the following questions:

a.  Describe how local richness ($\alpha$) and turnover ($\beta$) contribute to regional ($\gamma$) fish diversity in the Doubs.
b.  Is the fish assemblage at site 1 more similar to the one at site 2 or site 10?
c.  Using your understanding of the equation $\beta_{w} = \gamma/\alpha$, how would your interpretation of $\beta$ change if we instead defined beta additively (i.e., $\beta = \gamma - \alpha$)?

> ***Answer 3a***:  
> Answer 3b***: The fish assemblage at site 2 is more similar to site 1 than site 10.  
> Answer 3c***:  If beta was defined additively, you would get a lower beta value.  So, you would interpret beta as the difference between gamma and alpha.

### The Resemblance Matrix
In order to quantify $\beta$-diversity for more than two samples, we need to introduce a new primary ecological data structure: the **Resemblance Matrix**. 

***Question 4***: How do incidence- and abundance-based metrics differ in their treatment of rare species?

> Answer 4***:  Incidence-based metrics rely only on presence/absence which abundance-based metrics take into accound the number of individuals that are included in the sample.  This makes abundance-based metrics more quantitative; therefore they are able to account for the abundance of rare species elative to common species in the dataset.

In the R code chunk below, do the following:  

1. make a new object, `fish`, containing the fish abundance data for the Doubs River, 
2. remove any sites where no fish were observed (i.e., rows with sum of zero),
3. construct a resemblance matrix based on Sørensen's Similarity ("fish.ds"), and
4. construct a resemblance matrix based on Bray-Curtis Distance ("fish.db").  

```{r}
fish=doubs$fish
fish=fish[-8,]
fish.dj=vegdist(fish,method="jaccard",binary=T)
fish.db=vegdist(fish,method="bray")
fish.ds=vegdist(fish,method="bray",binary=T)
fish.db  #resemblance matrix Bray-Curtis based
fish.db=vegdist(fish,method="bray",upper=T,diag=T)  #square resemblance matrix
fish.db
fish.ds=vegdist(fish,method="bray",binary=T,upper=T,diag=T)
fish.ds
```

***Question 5***: Using the distance matrices from above, answer the following questions:

a. Does the resemblance matrix (`fish.db`) represent similarity or dissimilarity? What information in the resemblance matrix led you to arrive at your answer? 
b. Compare the resemblance matrices (`fish.db` or `fish.ds`) you just created. How does the choice of the Sørensen or Bray-Curtis distance influence your interpretation of site (dis)similarity?

> Answer 5a***:  This matrix represents dissimlarity.  You can deduce this from the relationship of a species to itself equalling zero.  This means that something is 0% different from itself (which makes perfect sense!). 
> Answer 5b***:  The abundance-based matrix seems to generate more dissimilarity.

## 4) VISUALIZING BETA-DIVERSITY

### A. Heatmaps

In the R code chunk below, do the following:  

1. define a color palette,
2. define the order of sites in the Doubs River, and
3. use the `levelplot()` function to create a heatmap of fish abundances in the Doubs River.

```{r}
order=rev(attr(fish.db, "Labels"))
levelplot(as.matrix(fish.db)[, order], aspect = "iso", col.regions = magma, 
          xlab = "Doubs Site", ylab = "Doubs Site", scales = list(cex = 0.5), 
          main = "Bray-Curtis Distance")
```

### B.  Cluster Analysis

In the R code chunk below, do the following:  

1. perform a cluster analysis using Ward's Clustering, and
2. plot your cluster analysis (use either `hclust` or `heatmap.2`). 

```{r} 
fish.ward=hclust(fish.db, method = "ward.D2")
par(mar = c(1, 5, 2, 2) + 0.1)
plot(fish.ward, main = "Doubs River Fish: Ward's Clustering",
     ylab = "Squared Bray-Curtis Distance")
gplots::heatmap.2(as.matrix(fish),distfun=function(x) vegdist(x,method="bray"),
        hclustfun=function(x) hclust (x,method="ward.D2"),
        col=viridis, trace="none",density.info="none")
```

***Question 6***: Based on cluster analyses and the introductory plots that we generated after loading the data, develop an ecological hypothesis for fish diversity the `doubs` data set? 

> Answer 6***: Due to some environmental factor, some species of fish prefer and/or are limited to particular parts of the river.  Along the river gradient there seems to be four clusterings, with one clustering being particularly different from the other three.

### C.  Ordination

### Principal Coordinates Analysis (PCoA)

In the R code chunk below, do the following:  

1. perform a Principal Coordinates Analysis to visualize beta-diversity
2. calculate the variation explained by the first three axes in your ordination
3. plot the PCoA ordination,
4. label the sites as points using the Doubs River site number, and
5. identify influential species and add species coordinates to PCoA plot.

```{r}
fish.pcoa=cmdscale(fish.db,eig=T, k=3)
explainvar1=round(fish.pcoa$eig[1] / sum(fish.pcoa$eig), 3) * 100
explainvar2=round(fish.pcoa$eig[2] / sum(fish.pcoa$eig), 3) * 100
explainvar3=round(fish.pcoa$eig[3] / sum(fish.pcoa$eig), 3) * 100
sum.eig=sum(explainvar1,explainvar2,explainvar3)

par(mar = c(5, 5, 1, 2) + 0.1)
plot(fish.pcoa$points[ ,1], fish.pcoa$points[ ,2], ylim = c(-0.2, 0.7),
    xlab = paste("PCoA 1 (", explainvar1, "%)", sep = ""),
    ylab = paste("PCoA 2 (", explainvar2, "%)", sep = ""),
    pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)
axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1) 
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1) 
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)
points(fish.pcoa$points[ ,1], fish.pcoa$points[ ,2], 
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(fish.pcoa$points[ ,1], fish.pcoa$points[ ,2], 
     labels = row.names(fish.pcoa$points))

fishREL=fish
  for(i in 1:nrow(fish)){
    fishREL[i, ] = fish[i, ] / sum(fish[i, ]) 
  }
fish.pcoa=add.spec.scores(fish.pcoa,fishREL,method = "pcoa.scores") 
text(fish.pcoa$cproj[ ,1], fish.pcoa$cproj[ ,2],
     labels = row.names(fish.pcoa$cproj), col = "black")
```

In the R code chunk below, do the following:  

1. identify influential species based on correlations along each PCoA axis (use a cutoff of 0.70), and
2. use a permutation test (999 permutations) to test the correlations of each species along each axis.

```{r}
spe.corr=add.spec.scores(fish.pcoa, fishREL, method = "cor.scores")$cproj
corrcut=0.7
imp.spp=spe.corr[abs(spe.corr[, 1]) >= corrcut | abs(spe.corr[, 2]) >= corrcut, ]

fit=envfit(fish.pcoa, fishREL, perm = 999)
```

***Question 7***: Address the following questions about the ordination results of the `doubs` data set:

a.  Describe the grouping of sites in the Doubs River based on fish community composition.
b.  Generate a hypothesis about which fish species are potential indicators of river quality.

> Answer 7a***:  There appears to be four site groupings based on community composition of fish, however one grouping consists of only few sites that are not tightly clustered.  The three clusters below the x-axis appear to have more sites for which the components are able to explain the variability in the community composition of fish.  However, PCoA axis 2 only explains 16.2% of the variabiliy in the grouping of sites based on species composition.
> Answer 7b***:   Phph, Rham, Legi, Blbj, and Alal are likely the most imporant indicators of river quality because they have a correlation-coeffeicient > 0.80, so changes in these species indepedently account for at least 80% of the change in species composition of fish.


## 5) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}

quality=c(rep("HQ", 13), rep("MQ", 5), rep("LQ", 6), rep("MQ", 5))

adonis(fish ~ quality, method = "bray", permutations = 999)

indval = multipatt(fish, cluster = quality, func = "IndVal.g", control = how(nperm=999)) 
summary(indval)

fish.rel=decostand(fish, method = "total")
phi = multipatt(fish.rel, cluster = quality, func = "r.g", control = how(nperm=999)) 
summary(phi)
```

***Question 8***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?

> Answer 8***: Habitat quality has a significant affect on fish species composition and quality account for approixmately 45.8% of fish species composition is explained by habitat quality.  However, individual species are grouped together differently depending on the type of test used (IndVal vs. phi coefficient).  In both cases, however, some fish are group together in a LQ/MQ.


### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to determine if these matrices are correlated, and test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
fish.dist=vegdist(doubs$fish[-8, ], method = "bray") 
env.dist=vegdist(scale(doubs$env[-8,]),method = "euclid")
mantel(fish.dist,env.dist)
```

***Question 9***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? How does this relate to your hypothesis about stream quality influencing fish communities?

> Answer 9***:  The Mantel test indicates that there is a moderately strong (Mantel r = 0.60) correlation between fish species composition and environmetnal variability.  This relates to the hypothesis about stream quality because it could mean that the type and abundance fo fish that change as a result of environmental variables may be related to the quality of the water.

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   

```{r}

env.chem=as.matrix(doubs$env[-8 , 5:11])

doubs.dbrda=dbrda(fish.db ~ ., as.data.frame(env.chem)) 
ordiplot(doubs.dbrda)

doubs.dbrda.mod0=dbrda(fish.db ~ 1, as.data.frame(env.chem))

ordiplot(doubs.dbrda.mod0)

doubs.dbrda.mod1=dbrda(fish.db ~ ., as.data.frame(env.chem))

doubs.dbrda=ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

doubs.dbrda$call 
doubs.dbrda$anova 
ordiplot(doubs.dbrda)

permutest(doubs.dbrda, permutations = 999) 
envfit(doubs.dbrda, env.chem[,c(4,6,7)], perm = 999)

dbrda.explainvar1=round(doubs.dbrda$CCA$eig[1] /
                          sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2=round(doubs.dbrda$CCA$eig[2] /
                          sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100

par(mar = c(5, 5, 4, 4) + 0.1)

plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1), ylim = c(-1.1, 2.7),
     xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)", sep = ""),
     ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

axis(side = 1, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1) 
axis(side = 2, labels = T, lwd.ticks = 2, cex.axis = 1.2, las = 1) 
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

points(scores(doubs.dbrda, display = "wa"),
       pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"),
     labels = row.names(scores(doubs.dbrda, display = "wa")))

vectors <- scores(doubs.dbrda, display = "bp")

arrows(0, 0, vectors[,1], vectors[, 2],
       lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[, 2], pos = 3, 
     labels = row.names(vectors))
axis(side = 3, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 1])) * 2, labels = pretty(range(vectors[, 1])))
axis(side = 4, lwd.ticks=2, cex.axis=1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[, 2])) * 2, labels = pretty(range(vectors[, 2])))
```

***Question 10***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> Answer 10***: Based on this ordination, biological oxygen demand, nitrate concentration, and dissolved oxygen are the most important environmetnal variables that exaplin fish species composition along the Doubs river.

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}
doubs.dbrda$anova
env.mod=model.matrix(~ oxy + bdo + nit, as.data.frame(env.chem))[,-1]

rs=rowSums(fish)/sum(fish)
doubs.pcnmw=pcnm(dist(doubs$xy[-8,]), w = rs, dist.ret = T)
doubs.pcnmw$values > 0

doubs.space=as.data.frame(scores(doubs.pcnmw))
doubs.pcnm.mod0=dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1=dbrda(fish.db ~ ., doubs.space)
step.pcnm=ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

plot(step.pcnm)

step.pcnm$anova

space.mod=model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 +
                            PCNM13 + PCNM16 + PCNM6, doubs.space)[,-1]

doubs.total.env=dbrda(fish.db ~ env.mod) 
doubs.total.space=dbrda(fish.db ~ space.mod)

doubs.env.cond.space=dbrda(fish.db ~ env.mod + Condition(space.mod)) 
doubs.space.cond.env=dbrda(fish.db ~ space.mod + Condition(env.mod))

permutest(doubs.env.cond.space, permutations = 999) 
permutest(doubs.space.cond.env, permutations = 999) 
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

doubs.varpart=varpart(fish.db, env.mod, space.mod) 
doubs.varpart

par(mar = c(2,2,2,2))
plot(doubs.varpart)
text(1, 0.25, "Space")
text(0, 0.25, "Env")
mtext("Variation Partitioning of\nDoubs Fish Diversity", side = 3, line = -3)

```

***Question 11***: Interpret the variation partitioning results.   

> Answer 11***: Spatial drivers account for 25% of the variation in fish community composition, environmental drivers account for the 13% of the variation in fish community composition, and the interaction between spatial and environmental drivers accounts for 37% of the variation in fish community composition.  Therefore, approximately 74% of the variation in fish community composition is exaplained by spatial and/or environmental drivers.

## SYNTHESIS
Load the dataset you are using for your project. Perform an ordination to visualize your dataset.  Using this ordination, develop some hypotheses relevant to $\beta$-diversity and identify the appropriate tools you would use to test them.